
\documentclass{article} % For LaTeX2e
\usepackage[submission]{colm2026_conference}

\usepackage{microtype}
\usepackage{hyperref}
\definecolor{darkblue}{rgb}{0, 0, 0.5}
\hypersetup{colorlinks=true, citecolor=darkblue, linkcolor=darkblue, urlcolor=darkblue}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsmath,amssymb} % math
\usepackage{xcolor}         % colors

\newcommand{\seth}[1]{{\color{olive}\textbf{Seth:} {#1}}}
\newcommand{\tersoo}[1]{{\color{red}\textbf{Tersoo:} #1}}
\newcommand{\joel}[1]{{\color{blue}\textbf{Joel:} #1}}
\newcommand{\kiran}[1]{{\color{green}\textbf{Kiran:} #1}}

% NOTE: including geometry package
% The geometery package modifies some page properties when used. This can dramatically change the page margins, leading to severe template violation, and potential desk rejection. If the package is required, it can be used with the "pass" flag to skip the default page modifications, as in the following line:
% \usepackage[pass]{geometry}

\usepackage{lineno}

\definecolor{darkblue}{rgb}{0, 0, 0.5}
\hypersetup{colorlinks=true, citecolor=darkblue, linkcolor=darkblue, urlcolor=darkblue}


% seth: take a look at a couple of these titles
\title{MetaHarness: An Automatic Reset-Free Self-Improving Harness for Autonomous VLM Agents}
% \title{}
% \title{MetaHarness: An Automatic Generalization of GeminiPlaysPokemon for Autonomous VLM Agents}
% \title{AutoEvolve: A Reset-Free Evolutionary Self-Improving Harness for Autonomous Agents}

% Authors must not appear in the submitted version. This should be be taken care of automatically as long as you are using the "submission" option for the colm2026_conference package. But it's on the authors to verify. Non-anonymous submissions will be rejected without review.

\author{
Seth Karten*  \and
Joel Zhang* \and
Tersoo Upaa Jr* \and
Ruirong Feng \and
Wenzhe Li \and
Chengshuai Shi \and
Chi Jin \and
Kiran Vodrahalli
}

% The \author macro works with any number of authors. There are two commands
% used to separate the names and addresses of multiple authors: \And and \AND.
%
% Using \And between authors leaves it to \LaTeX{} to determine where to break
% the lines. Using \AND forces a linebreak at that point. So, if \LaTeX{}
% puts 3 of 4 authors names on the first line, and the last on the second
% line, try using \AND instead of \And before the third author name.

\newcommand{\fix}{\marginpar{FIX}}
\newcommand{\new}{\marginpar{NEW}}

\begin{document}

\ifcolmsubmission
\linenumbers
\fi

\maketitle

\begin{abstract}
Agentic harnesses, the scaffolding that wraps a foundation model with tools, memory, and planning, have become standard infrastructure for autonomous agents. In coding domains, harnesses such as Claude Code and OpenHands provide structured environments for software engineering tasks, but no analogous harness exists for embodied agents, where long-horizon decision-making under partial observability poses distinct challenges. We introduce \textbf{MetaHarness} \kiran{I would actually avoid calling it evolution, it's not really about evolving - it's more like in-context online RL}, a self-improving harness for embodied agents that requires only a minimal environment interface (minimal map info, frame observations and button inputs) and constructs its own scaffolding, including sub-agents, memory systems, and skill libraries, through in-context evolutionary optimization. Unlike prior prompt-evolution methods that require environment resets between iterations, MetaHarness operates \emph{reset-free}: it evolves its own system prompt, skill prompts, and sub-agent structure using trajectory segments gathered during a single continuous episode. We first present results from the Gemini Plays Pokémon (GPP) project, a series of long-running experiments in which we iteratively evolved a harness with human-in-the-loop feedback, producing the first AI system to complete multiple Pokémon RPGs. GPP agents beat Pokémon Blue, defeated the Elite Four in Pokémon Yellow Legacy on hard mode (with level caps and set battle style), and completed Pokémon Crystal without losing a single battle, spontaneously developing complex strategies such as multi-stage stalling plans and self-constructed battle sub-agents along the way. We then formalize and automate this process with MetaHarness. We evaluate on two Pokémon RPGs (Red and Emerald), comparing against a minimalist baseline and the human-designed expert harness from PokeAgent across multiple frontier models. As a proof of concept, we show that evolved harnesses transfer to open-source models via fine-tuning. MetaHarness improves during gameplay, approaches the performance of human-engineered harnesses, and enables faster re-traversal of previously encountered game segments with its self-updated scaffolding.
\end{abstract}

\input{sections/01_introduction}
\input{sections/02_preliminaries}
\input{sections/03_methodology}
\input{sections/04_experiments}
\input{sections/05_related_work}
\input{sections/06_discussion}

% \section*{Author Contributions}
% Seth, Joel, Tersoo co-first
% If you'd like to, you may include  a section for author contributions as is done
% in many journals. This is optional and at the discretion of the authors.

% \section*{Acknowledgments}
% Use unnumbered first level headings for the acknowledgments. All
% acknowledgments, including those to funding agencies, go at the end of the paper.


\bibliography{root}
\bibliographystyle{colm2026_conference}

% \appendix
% \section{Appendix}
% You may include other additional sections here.

\end{document}
