% ============================================================================
% Chapter 2, Section 1: Introduction
% ============================================================================

\section{Background}\label{sec:background}

\subsection{Introduction}\label{sec:introduction}

The field of Artificial Intelligence (AI) is undergoing a fundamental shift. Where language models were once evaluated on their ability to produce fluent text in response to short prompts, the frontier has moved toward sustained, autonomous interaction with complex environments. Systems are now expected to write and debug code across entire repositories, operate robotic platforms, and navigate virtual worlds---tasks that demand not just knowledge but the ability to plan, adapt, and maintain coherence over extended horizons.

Partial observability, long-horizon planning, and multi-modal perception are fundamental problems at the core of these emerging workflows. In these settings, agents are not only required to iterate on decisions that seek to optimize the successful completion of near-horizon tasks, but must also effectively plan for and maintain coherence across many thousands of environment interactions in pursuit of larger, often more abstract goals. In light of this new paradigm, there have been recent efforts to test the capabilities of frontier Large Language Models (LLMs) and the agent systems they underpin---many of these efforts finding that powerful Harness-Layer abstractions are just as, if not more central to the success of AI systems in these settings. Yet few existing benchmarks stress all three dimensions simultaneously under realistic conditions. Standard testbeds tend to isolate one axis: imperfect-information games emphasize equilibrium computation in short episodes, while open-ended environments test exploration but lack adversarial opponents or extended temporal commitments. As frontier language and vision--language models (VLMs) are increasingly deployed as autonomous agents to navigate codebases, serve as the reasoning layer for robotic systems, and interact with virtual environments that require multi-modal grounding, the gap between what existing benchmarks measure and what real-world autonomy demands has grown acute.

Pok\'{e}mon role-playing games (RPGs) offer a compelling case-study for this exact purpose. They combine a vast, partially observable state space with long-horizon, often non-linear narrative progression that demands thousands of cumulative decisions spanning exploration, resource management, combat, and spatial navigation. The recent surge of interest in Pok\'{e}mon-playing AI systems---Claude Plays Pok\'{e}mon~\citep{anthropic2025visible}, Gemini Plays Pok\'{e}mon~\citep{zhang2025geminiplayspokemon}, and GPT-5's completion of Pok\'{e}mon Red~\citep{engadget2025gptplayspokemon}---has demonstrated both the appeal of this domain and its limitations as a research vehicle. Fragmented across different games, harnesses, and evaluation criteria, the disconnected nature of these projects has made it difficult to rigorously disentangle both quantitative and qualitative metrics for success from the performance of the underlying model, sophistication of the harness architecture, or hardcoded assumptions about access to environment state.

This work addresses the fragmentation problem, advancing the state of the art across three complementary dimensions:

We first introduce an open \textbf{speedrunning benchmark}, as a robust foundation for evaluating frontier model and harness capabilities in long-duration, embodied settings and builds upon the testing framework established by the Speedrunning Track of the \pokeagent{} Challenge at NeurIPS 2025~\citep{karten2025pokeagent}. Our benchmark formalizes Pok\'{e}mon Emerald gameplay as a structured evaluation with standardized milestones, quantitative metrics, and infrastructure for fair cross-paradigm comparison. We have since extended it from the tutorial segment (through the first gym) to a long-horizon evaluation spanning many more segments of the game, where experiments proceed on the order of days and tens of thousands of environment interactions. Next, we present \textit{PokeAgent,} a state-of-the-art (SOTA) \textbf{custom VLM-agent harness} tasked with maintaining long-context coherence, developing and iterating upon plans, and effective problem solving across the Pok\'{e}mon Emerald storyline. Here, we detail the many insights spanning the development of PokeAgent from a simple four-module scaffold during the NeurIPS competition into a multi-agent system with powerful abstractions for processing global session-level context, hierarchical memory management, skill-learning, subagent orchestration, and sandboxed code execution. We further compare PokeAgent against common CLI-agent harnesses (Claude Code, Codex CLI, Gemini CLI) revealing that, while coding-agent architectures are impressive out-of-the-box, they
nonetheless struggle to properly adapt to the unique set of constraints imposed by our speedrunning benchmark---a result that underscores the benchmark's capacity to discriminate between architectures along axes of performance that general-purpose agent designs do not yet address. Finally, we introduce \textbf{AutoEvolve}, a meta-harness framework and reset-free evolutionary algorithm designed to automate experiential learning by iteratively building agent scaffolding through in-context optimization over global trajectory segment history. Beginning from the simplest initialization, AutoEvolve seeks to bootstrap the development of its own harness by pruning and refining memories, distilling agent progress across entire episodes of environment interactions into skill and subagent abstractions, and refining system prompts accordingly with respect to the evolving harness. Unlike prior methods for model-level and agent self-adaptation that require full episodes and environment resets~\citep{agrawal2025gepa}, AutoEvolve operates in a fully online, reset-free manner---targeting continuous self-improvement during a single extended gameplay session. We present preliminary results and analysis in Chapter~5.

\tersoo{note: an important note for cohesion, if i introduce a term like Embodied Agent Environment and Agentic Harness, I should be more intentional in using it in place of more generic terms when applicable... especially if i give it an acronym}
The remainder of this work is organized as follows. The present chapter continues with necessary motivation and context (Section~\ref{sec:motivation}), surveying related research across embodied agents, self-adaptation, and the emerging field of meta-harnesses (Section~\ref{sec:related_work}), and formalizes our conception of \textbf{Embodied Agent Environments (EAEs)} and \textbf{Agentic Harnesses}, including a spectrum of harness configurations that frames the experimental comparisons in later chapters (Section~\ref{sec:formalization}). Chapter~\ref{sec:benchmark} describes the speedrunning benchmark in detail: its environment design, evaluation criteria, and baselines. Chapter~\ref{sec:pokeagent} traces the development of the PokeAgent harness from its competition-era origins through its current multi-agent architecture, presenting early performance results that motivate the need for automated harness evolution. Chapter~\ref{sec:autoevolve} presents AutoEvolve, including its methodology, experimental design across many agent-evaluation settings, and a discussion of our results. Chapter~\ref{sec:outlook} concludes with an exploration of the implications of this work and fruitful areas for future investigation across model distillation, ultra-long context (ULC) evaluations, and extensions to domains beyond Pok\'{e}mon.
