% ============================================================================
% Chapter 2, Section 3: Related Work
% ============================================================================

\subsection{Related Work}\label{sec:related_work}

We survey several lines of research that collectively inform the design of our benchmark, harness, and AutoEvolve algorithm: reinforcement learning, experiential learning, recursive language models, long-horizon benchmarks, embodied agents in RPG settings, self-adaptation methods, and the emerging field of meta-harnesses.

\subsubsection{Reinforcement Learning}\label{sec:rl}

Reinforcement learning (RL) provides the classical framework for sequential decision-making under uncertainty. An agent interacts with an environment modeled as a Markov Decision Process (MDP), selecting actions to maximize cumulative reward over time. When the full environment state is not observable, the problem generalizes to a Partially Observable MDP (POMDP), requiring the agent to maintain beliefs over hidden states---a setting that closely mirrors the challenges of Pok\'{e}mon gameplay.

RL has produced landmark results in game AI. Deep Q-Networks~\citep{mnih2015human} achieved human-level performance on Atari games by combining deep neural networks with experience replay. AlphaGo and its successors~\citep{silver2017mastering, silver2018general} mastered Go, chess, and shogi through self-play, while AlphaStar~\citep{vinyals2019grandmaster} reached grandmaster level in StarCraft~II using multi-agent RL. These results demonstrate RL's capacity for superhuman play---but principally in settings that are fully observable, deterministic, or narrow enough for exhaustive exploration.

A consistent pattern emerges as environments grow more complex: RL's margin over human performance erodes under partial observability, stochastic dynamics, and enormous state spaces~\citep{samuel1959, paglieri2024balrog}. In Pok\'{e}mon, the state space is so vast ($\sim\!10^{564}$ battle states alone) and the observation so partial that pure RL approaches have yet to prove scalable and let alone achieve competitive performance without significant domain engineering~\citep{pleines2025pokemon, grigsby2025human}.

LLM-based agents represent a fundamentally different paradigm. Rather than learning policies from scratch through reward signals, they leverage vast pretrained world knowledge of game mechanics, strategies, and narrative structure, and apply it through in-context reasoning. This trades RL's sample efficiency and formal optimality guarantees for breadth of knowledge and zero-shot generalization. The tension between the paradigms of structured optimization central to RL versus reliance on pretrained knowledge and in-context LLM reasoning is one that remains central to the development todays AI systems \tersoo{cite!}.

\subsubsection{Experiential Learning}\label{sec:experiential_learning}

A growing body of work explores how agents can accumulate and leverage experience over extended interactions, moving beyond static policy deployment toward systems that learn through doing. While experiential learning has a deep tradition in RL paradigm---often through self-play--- where policies are optimized through reward signals over many episodes, experiential learning in the LLM setting emphasizes structured knowledge accumulation within and across sessions, using the model's own reasoning and reflection as the primary learning signal.

AriGraph~\citep{anokhin2024arigraph} constructs knowledge graph world models with episodic memory for LLM agents, enabling them to maintain structured representations of their environment across interactions. This approach addresses a key limitation of standard LLM agents: the inability to retain and organize discoveries, spatial relationships, and causal knowledge beyond the current context window. By externalizing knowledge into a graph structure, AriGraph demonstrates that persistent, structured memory can compensate for the finite context window and enable agents to reason over accumulated experience rather than raw interaction logs.

GRACE~\citep{song2024grace} introduces generalist agents that leverage both code generation and experiential learning, demonstrating that combining procedural knowledge (executable code) with episodic knowledge (accumulated experience) produces more robust agent behavior than either alone. This dual-knowledge paradigm that seeks both executable routines and behavioral heuristics has emerged as a design pattern in several subsequent agent architectures. \tersoo{note: edit this paragrpah to maintain focus}

Language Models Meet World Models~\citep{xiang2023embodied} showed that grounding LLMs in embodied experience rather than relying solely on pretraining knowledge, significantly improves their performance on downstream reasoning tasks. In this setting, structured experiential data serves as a powerful complement to pretrained knowledge, enabling agents to adapt to novel environments that diverge greatly from their training distribution.

Recent work on in-context reinforcement learning (ICRL) has demonstrated that LLMs can perform implicit policy improvement from in-context reward signals~\citep{song2025icrl, karten2025economist}. As we will see, if language models can implicitly improve policies from reward signals presented in-context, then it holds to reason that explicitly grounding these signals in the agent's trajectory and reflection process may complement this. It is this principle that informs several of the methods discussed in Chapter 5. \tersoo{NOTE: this is a powerful sentence. As a matter of continuity, definitely make sure to revist this line of reasoning in chapter 5.}

\subsubsubsection{\textbf{Skill Learning.}} A recurring theme in experiential learning is the automatic discovery and curation of reusable behavioral skills. Voyager~\citep{wang2023voyager} demonstrated that language model agents can progressively acquire capabilities through interaction, storing successful programs in a growing code library indexed by natural language descriptions. GRACE~\citep{song2024grace} extended this with dual code-and-experience skills, showing that procedural and episodic knowledge are complementary.

SkillRL~\citep{xia2026skillrl} provides the most comprehensive treatment of this problem to date. Rather than simply accumulating raw, often noise and redundant trajectories, SkillRL introduces an experience-based distillation mechanism that extracts high-level behavioral patterns into a hierarchical skill library (SkillBank). An adaptive retrieval strategy enables the agent to access both general and task-specific heuristics at inference time, while a recursive evolution mechanism allows the skill library to co-evolve with the agent's policy during reinforcement learning training. On the ALFWorld, WebShop, and seven search-augmented benchmarks, SkillRL achieves state-of-the-art performance, outperforming strong baselines by over 15.3\% while significantly reducing the token footprint. \tersoo{NOTE: this is a powerful discussion. As a matter of continuity, definitely make sure to revist this line of reasoning in chapter 4/5.}

\subsubsubsection{\textbf{Agent Memory.}} Memory is the mechanism by which experiential learning persists beyond the immediate context window, and its design has emerged as a critical architectural decision for long-horizon agents. Agent memory systems generally organize knowledge along three axes: \emph{episodic memory} (logs of specific interactions and events), \emph{semantic memory} (accumulated facts, strategies, and world knowledge), and \emph{procedural memory} (learned skills and behavioral routines).

Generative Agents~\citep{park2023generative} introduced an influential memory architecture that combines recency, importance, and relevance-based retrieval to select which memories to surface during reasoning. This design enables believable long-term behavior in simulated social environments and has influenced subsequent work on agent memory, including retrieval-augmented approaches that treat memory as an external knowledge base queried at each reasoning step.

MemGPT~\citep{packer2023memgpt} draws an explicit analogy between LLM context management and operating-system virtual memory. By implementing a hierarchical memory system with fast (in-context) and slow (external storage) tiers, MemGPT enables agents to intelligently page information in and out of their context window, maintaining coherent behavior across interactions that would otherwise exceed context limits. This approach is particularly relevant to the long-context settings we consider, where agents must maintain coherence across tens of thousands of environment interactions by using memory abstractions to localize their progress within a more long-term objective.

AriGraph~\citep{anokhin2024arigraph}, discussed above, takes a complementary approach by structuring memory as a knowledge graph with episodic annotations. Together, these works establish that effective memory design is deeply invested in the nuances of how experience is stored, organized, retrieved, and eventually deprecated, and is as important to agent performance as the latent reasoning capabilities of the foundation model.

\subsubsection{Recursive Language Models}\label{sec:rlm}

Recursive Language Models (RLMs)~\citep{zhang2025rlm} formalize the paradigm of LLMs that programmatically interact with and recurse over their own context. Rather than treating the context window as a passive repository of information, RLMs treat it as an active workspace that the model can read from, write to, and restructure during inference. This provides a theoretical foundation for agents that self-modify their reasoning process---moving beyond simple chain-of-thought reasoning toward systems that can dynamically manage their entire session history.

The RLM framework is particularly relevant to the problem of \emph{context rot}: the empirically observed degradation of coherent reasoning as context windows grow large. As agents accumulate interaction history, the ratio of relevant to irrelevant information in-context deteriorates, leading to repetitive behavior, lost goals, and contradictory reasoning. These failure modes are documented across long-horizon agent systems from Claude Plays Pok\'{e}mon~\citep{anthropic2025visible} to Gemini Plays Pok\'{e}mon~\citep{zhang2025geminiplayspokemon}. By treating context as a structured, programmatically accessible workspace rather than a flat append-only log, RLMs suggest an architectural principle for combating this degradation.

\tersoo{NOTE: this is a powerful discussion. As a matter of continuity, definitely make sure to revist this line of reasoning in chapter 4/5.... The RLM framework directly informs the design of both our expert harness (Chapter~4), which maintains a structured trajectory window that subagents process and summarize, and our AutoEvolve algorithm (Chapter~5), which can be understood as a form of recursive self-modification operating over trajectory segments. We defer the detailed connections to those chapters. } 

\subsubsection{Long-Horizon Benchmarks}\label{sec:long_horizon_benchmarks}

Concurrent with our work, several benchmarks have emerged that specifically target long-horizon agent evaluation, independently validating that this capability axis is underserved by existing evaluation suites.

DeepPlanning~\citep{zhang2026deepplanning} benchmarks long-horizon agentic planning with verifiable constraints, requiring agents to perform multi-day travel planning and multi-product shopping scenarios that demand proactive information acquisition, local constrained reasoning, and global budget optimization. Even frontier models struggle with these tasks, highlighting the gap between short-context reasoning benchmarks and real-world planning demands.

APEX-Agents~\citep{vidgen2026apex} evaluates whether AI agents can execute long-horizon, cross-application tasks drawn from investment banking, management consulting, and corporate law. Agents must navigate realistic work environments with files and tools; the best-performing system (Gemini 3 Flash) achieves only 24.0\% on the benchmark, confirming that long-horizon task completion in professional settings remains far from solved.

HeroBench~\citep{anokhin2025herobench} is closest in spirit to our own benchmark. It evaluates long-horizon planning and structured reasoning in RPG-inspired virtual worlds, where tasks require strategic planning, resource gathering, skill mastery, equipment crafting, and combat---reflecting the layered dependencies characteristic of role-playing games. An evaluation of 25 state-of-the-art LLMs reveals substantial performance disparities rarely seen in conventional reasoning benchmarks, with detailed error analysis exposing weaknesses in high-level plan generation and structured action execution.

These benchmarks collectively reinforce the notion that long-horizon, embodied evaluation represents a genuinely distinct axis of model competence. Our speedrunning benchmark complements these efforts by centering visual perception, partial observability, and multi-hour continuous operation in a living game environment.

\subsubsection{Embodied Agents in RPG Settings}\label{sec:embodied_agents}

The deployment of language models as controllers for embodied agents has produced a rich body of work exploring how latent world knowledge can be translated into effective physical or virtual action. Voyager~\citep{wang2023voyager} demonstrated open-ended skill acquisition in Minecraft through an automatic curriculum and a growing code library, achieving lifelong learning without manual reward engineering. However, Voyager relies on a programmatic API that provides structured state information (inventory contents, nearby blocks, entity positions) without access to  raw perceptual input, sidestepping the perception challenges central to RPG gameplay and decreasing the generalizeability of its method to non-Minecraft domains.

SayCan~\citep{brohan2023can} grounded language model instructions in robotic affordances, showing that LLMs can select contextually appropriate actions when paired with learned value functions---but operates in short-horizon manipulation tasks rather than extended autonomous episodes. Code as Policies~\citep{liang2023code} generalized this approach by having language models generate executable robot control code, demonstrating compositional generalization to novel instructions. GITM (Ghost in the Minecraft)~\citep{zhu2023ghost} combined LLMs with text-based knowledge and memory to create generally capable open-world agents, introducing a knowledge graph for tracking world state. Steve-Eye~\citep{zheng2024steveeye} extended this line of work by equipping LLM-based agents with visual perception in Minecraft, moving closer to the perceptual challenges we face in Pok\'{e}mon.

Generative Agents~\citep{park2023generative} demonstrated that LLMs can produce believable human behavior in a simulated town environment when equipped with memory retrieval, reflection, and planning mechanisms. Their memory architecture---combining recency, importance, and relevance-based retrieval---has influenced subsequent work on agent memory systems, including our own hierarchical memory design in PokeAgent.

A common thread across these works is that none operates from purely perceptual input (raw frames) in a long-horizon RPG setting requiring sustained reasoning over many hours. The closest analogue to our setting is Voyager, but it benefits from a structured API, operates in episodes of minutes rather than hours, and does not contend with the partially observable, narratively structured progression that characterizes Pok\'{e}mon RPGs. Our work bridges this gap by developing a harness that processes raw visual observations alongside structured state text and supports multi-hour continuous operation.

\subsubsection{Self-Adaptive Methods}\label{sec:self_adaptation}

Several lines of work explore how language models and LLM-based agents can improve their own performance through explicit abstractions for self-directed learning, each making different tradeoffs between the scope of adaptation, the requirement for environment resets, and the computational overhead involved.

\paragraph{Prompt Evolution.} GEPA~\citep{agrawal2025gepa} uses genetic evolution over prompt candidates, guided by natural-language reflections on agent trajectories. It runs an agent multiple times, collects reasoning traces and outcomes, and has the model analyze these traces to propose prompt edits. Candidates that improve performance are retained for the next generation. GEPA demonstrated that this in-context, reflective evolution can outperform RL-based fine-tuning across several benchmarks, establishing prompt evolution as a viable alternative to weight updates. The method's primary limitation, however, is its requirement of full episodic completion and environment resets between iterations: each evolution step begins from the initial state, making it less directly applicable to long-context RL settings like the Pok\'emon where resets are expensive or unavailable.

MIPROv2~\citep{opsahlong2024miprov2} applies multi-stage prompt optimization to DSPy programs, optimizing both instructions and demonstrations through Bayesian surrogate models. While effective for modular NLP pipelines, it assumes a fixed program structure and does not extend to the open-ended harness evolution that AutoEvolve targets.

\paragraph{Weight-Level Self-Adaptation.} Self-Adapting LLMs (SEAL)~\citep{zweiger2025seal} take an RL-first approach, enabling LLMs to generate their own fine-tuning data and apply policy-gradient updates to themselves. Over time, the model's weights evolve to incorporate new knowledge. While SEAL yields persistent improvements, it requires careful training to avoid catastrophic forgetting and operates at the weight level rather than the scaffold level. Our approach leaves model weights entirely unchanged and instead adapts via harness optimization---a strictly inference-time strategy that is simpler to deploy and leaves the base model available for other tasks.

\paragraph{Self-Refine.} The Self-Refine technique~\citep{madaan2023selfrefine} formalized the intuitive idea of iterative self-critique: the model produces an initial solution, generates feedback highlighting flaws, and revises its solution accordingly. This loop can be repeated to elicit higher-quality results without external feedback. Self-Refine operates within a single inference call and does not persist improvements across interactions. Our reflection and subagent mechanisms in PokeAgent (Chapter~4) extend this idea to the agent setting, where a dedicated reflecting subagent critiques the orchestrator's recent behavior and produces structured feedback that informs subsequent decisions.

\paragraph{Reset-Free Optimization.} A key distinction of our work is the \emph{reset-free} setting. Standard prompt evolution~\citep{agrawal2025gepa} and most RL-based methods require resetting the environment to a known initial state between optimization iterations. In the long-horizon RPG setting, this is problematic: a single speedrunning episode can last 12+ hours, and progress is inherently cumulative (items collected, battles won, areas unlocked). AutoEvolve evolves its harness mid-episode using trajectory \emph{segments} rather than complete episodes, enabling continuous adaptation without interrupting the agent's progress. This reset-free property distinguishes AutoEvolve from all prior prompt-evolution methods and is central to its applicability in embodied settings.

\paragraph{LLM-as-a-Judge.} The use of language models as automated evaluators has become widespread, with MT-Bench and Chatbot Arena~\citep{zheng2023mtbench} establishing the paradigm of LLM-as-a-Judge for open-ended evaluation. Subsequent work has identified systematic biases in this approach, including position bias and verbosity bias~\citep{zhao2024llmjudgefailures}, and has extended the paradigm to multimodal settings~\citep{chen2024mllmjudge}. AutoEvolve's evolutionary optimization relies on an LLM judge module that evaluates trajectory segments and produces structured critiques. We address known biases through careful prompt design and by grounding the judge's evaluations in concrete game-state metrics (milestone progress, step counts, spatial positioning) rather than relying solely on qualitative assessment.

\subsubsection{Meta-Harnesses}\label{sec:meta_harnesses}

There is a growing body of work which seeks to optimize and automate the creation of the agent harness itself, including its constituent system prompts, tool definitions, retry logic, context management, and subagent configuration that wrap a foundation model as an optimizable system. This emerging research direction recognizes that harness quality has become a first-order variable in agent performance, often mattering more than the choice of underlying model---a claim supported empirically by Meta-Harness's demonstration of a 6$\times$ performance gap through harness optimization alone~\citep{lee2026metaharness} and AutoHarness's finding that a smaller model with a synthesized harness outperforms larger models without one~\citep{lou2026autoharness}.

\paragraph{The Darwin G\"{o}del Machine (DGM).} \citet{zhang2025dgm} introduced open-ended evolution of self-improving coding agents. DGM maintains an archive of generated coding agents and grows the archive by sampling an agent and using a foundation model to create new, ``interesting'' variants---inspired by quality-diversity algorithms from open-endedness research. This process produces a growing tree of diverse, high-quality agents. Empirically, DGM improved coding capabilities on SWE-bench from 20.0\% to 50.0\% and on Polyglot from 14.2\% to 30.7\%, discovering improvements such as better code editing tools, long-context window management, and peer-review mechanisms. DGM operates across episodes (each agent variant is evaluated independently), requiring environment resets between iterations.

\paragraph{Meta-Harness.} \citet{lee2026metaharness} automated harness optimization by giving a coding agent (Claude Code) access to a filesystem containing the full source code, execution traces, and performance scores of all prior harness candidates---up to 10 million tokens of diagnostic context per optimization step. This 400$\times$ increase over prior methods (which typically provided at most 26K tokens of summary information) enables the optimizer to trace failures to specific lines of harness code or prompt phrases. Meta-Harness achieved 76.4\% on Terminal-Bench 2.0 with Claude Opus 4.6 and demonstrated a 6$\times$ performance gap on the same underlying model through harness optimization alone.

\paragraph{AutoHarness.} \citet{lou2026autoharness} demonstrated that Gemini-2.5-Flash can automatically synthesize code harnesses through iterative refinement guided by environment feedback, using tree search with Thompson sampling to explore the space of potential harnesses. The resulting harnesses prevent illegal moves in 145 TextArena games, enabling the smaller Gemini-2.5-Flash to outperform larger models including Gemini-2.5-Pro and GPT-5.2-High. AutoHarness focuses on \emph{constraint enforcement}---preventing illegal actions---rather than the broader harness evolution (prompts, subagents, skills, memory) that AutoEvolve targets.

Our AutoEvolve framework differs from all three approaches in several key respects. First, AutoEvolve is \emph{fully online and reset-free}: it evolves the harness during a single continuous episode rather than across independent evaluation runs. Second, it is \emph{tailored for embodied settings} with raw perceptual input (frame observations and button inputs), whereas DGM, Meta-Harness, and AutoHarness operate in textual environments with structured programmatic feedback. Third, AutoEvolve evolves the \emph{complete harness}---system prompt, subagent registry, skill library, and memory store---through four coordinated evolutionary passes, rather than optimizing a single component (code, prompts, or constraints) in isolation. These differences reflect the distinct demands of long-horizon embodied autonomy, where the agent cannot restart, perception is noisy, and the space of useful scaffolding is too large for any single-axis optimization to explore effectively.
