% ============================================================================
% Chapter 3: Speedrunning Benchmark (Headers Only)
% ============================================================================

\newpage
\section{Speedrunning Benchmark}\label{sec:benchmark}

\subsection{NeurIPS Competition Background}\label{sec:neurips_background}

\textit{[To be written. The \pokeagent{} Challenge as a NeurIPS 2025 competition; evolution to a living benchmark extended through three gyms.]}

\subsection{Environment Design}\label{sec:environment_design}

\subsubsection{The World from the Perspective of the Agent}\label{sec:agent_perspective}

\textit{[To be written. Key preliminaries: observation space, action space, game mechanics visible to the agent.]}

\subsubsection{Tutorial}\label{sec:tutorial}

\textit{[To be written. Tests essential components; first gym (Roxanne); simplest setting that must pass reliably before long-context ambition. Figure from benchmark paper.]}

\subsubsection{Long Horizon}\label{sec:long_horizon}

\textit{[To be written. Extended evaluation through the third gym; experiments exceeding 12+ hours; tests long-context coherence, decision making, non-linear progression, problem solving. New figure based on benchmark paper.]}

\subsection{Speedrunning Evaluation Criteria}\label{sec:evaluation_criteria}

\subsubsection{Quantitative Metrics}\label{sec:quantitative_metrics}

\textit{[To be written. Milestone completion percentage, wall-clock time, steps per milestone, token usage.]}

\subsubsection{Qualitative Analysis}\label{sec:qualitative_analysis}

\textit{[To be written. Less structured but richer analysis of agent behavior nuances.]}

\subsection{Speedrunning Baselines}\label{sec:baselines}

\subsubsection{Human Baselines}\label{sec:human_baselines}

\textit{[To be written. Human speedrun reference times.]}

\subsubsection{Harness vs.\ Model Capability}\label{sec:harness_vs_model}

\textit{[To be written. Disentangling harness contribution from model contribution; five-dimensional characterization (S, T, M, F, $\Phi$).]}
