% ============================================================================
% Chapter 5: AutoEvolve (Headers Only)
% ============================================================================

\newpage
\section{AutoEvolve}\label{sec:autoevolve}

\subsection{Beyond Pok\'{e}mon}\label{sec:beyond_pokemon}

\tersoo{NOTE: We will need to have a table figure here with the specifics around what types of tools each H_min, H_expert, H_auto has access to has access to!}

\tersoo{NOTE: this section will pull heavily from the MetaHarness paper draft for NeurIPS 2026... its basically a mini-paper on its own, without deep discussion into related work.}
\textit{[To be written. Position AutoEvolve as a domain-agnostic evolutionary loop requiring modular agent framework, sandboxing, and a global session log; leverages LLM-as-a-judge in a reset-free manner. AutoEvolve is inherently domain agnostic.]}

\subsection{Methodology}\label{sec:methodology}

\textit{[NOTE from Ch.~2 refactor: The following details were moved here from the Formalization and Related Work sections of Chapter~2, to be integrated into the methodology.]}

\tersoo{--- Content moved from Chapter 4 ---
 These paragraphs discuss how H_expert abstractions become the substrate for AutoEvolve.
 Integrate into the methodology narrative when writing this section.

 (1) Trajectory window as bridge between H_expert and H_auto:
 The trajectory window provides the primitive that becomes central to $\mathcal{H}_{\min}$ and $\mathcal{H}_{\text{auto}}$. In $\mathcal{H}_{\text{expert}}$, the trajectory is processed by built-in subagents with hand-crafted prompts. In $\mathcal{H}_{\text{auto}}$, the same trajectory processing infrastructure is exposed to the evolutionary loop, which can create custom subagents that operate on trajectory segments in novel ways. The trajectory window thus serves as the bridge between the human-engineered and the auto-evolved harness configurations.

 (2) Skill library population in H_expert vs H_auto:
 In $\mathcal{H}_{\text{expert}}$, the skill library is typically populated by the agent during gameplay as it identifies recurring patterns. In $\mathcal{H}_{\text{auto}}$, the AutoEvolve loop can also write skills, enabling the evolutionary process to encode discovered strategies as reusable code.

 (3) Custom subagent primitive and AutoEvolve expressivity:
 The general subagent primitive increases the generality of the harness and the expressivity of $\mathcal{H}_{\text{auto}}$: the AutoEvolve loop can define new subagents with novel system instructions, tool sets, and termination conditions, effectively extending the agent's behavioral repertoire without human engineering.}

\textit{[RLM connection: The trajectory window in PokeAgent functions as an RLM-style workspace---the full history of the agent's execution (state observations, prompts, reasoning traces, actions) is maintained as a structured, programmatically accessible log that subagents process and summarize for the orchestrator. This is a departure from standard agent architectures that treat history as a flat, append-only sequence. AutoEvolve's mid-episode evolution can be understood as a form of recursive self-modification: the agent analyzes its own trajectory segments and uses this analysis to rewrite its harness components (system prompt, subagents, skills, memory), which in turn shape subsequent trajectory segments. Unlike the depth-$k$ recursion that RLMs formalize, AutoEvolve operates at depth-1.]}

\textit{[Harness spectrum: Define $\mathcal{H}_{\min}$ (minimalist), $\mathcal{H}_{\text{expert}}$ (human-designed), $\mathcal{H}_{\text{meta}}$ (meta-tool), $\mathcal{H}_{\text{auto}}$ (auto-evolved). Central question: whether $\mathcal{H}_{\text{auto}}$, evolved from $\mathcal{H}_{\min}$ through AutoEvolve's reset-free optimization, can approach the performance of $\mathcal{H}_{\text{expert}}$.]}

\textit{[In-context prompt evolution formalism: episodic setting (Eq.~for GEPA) vs. reset-free setting (Eq.~for AutoEvolve). Four-pass structure over prompts, subagents, skills, and memory.]}

\subsubsection{Algorithm}\label{sec:algorithm}

\textit{[To be written. Four-pass evolutionary optimization: system prompt, subagent registry, skill library, memory store. Scheduling (warmup steps, evolution frequency). Detailed pseudocode.]}

\subsubsection{Experimental Design}\label{sec:experimental_design}

\textit{[To be written. Three experimental settings, each with three trials:]}

\begin{enumerate}
    \item \textit{Three scaffold tiers: Baseline (simple) / Baseline + AutoEvolve / Fully Custom PokeAgent}
    \begin{itemize}
        \item \textit{Test Gemini 3.1-pro-preview}
        \item \textit{Test Gemini 3-flash-preview}
    \end{itemize}
    \item \textit{Sophisticated Supervisor: powerful model for evolution, weaker model for orchestrator}
    \item \textit{Bootstrapped Transfer Learning: equip weaker orchestrator with auto-evolved harness from stronger model + supervisor}
\end{enumerate}

\subsubsection{Metrics}\label{sec:autoevolve_metrics}

\textit{[To be written. Standard speedrunning evaluation criteria; qualitative observations for each setting.]}

\subsection{Results}\label{sec:results}

\subsubsection{Three Settings}\label{sec:three_settings_results}

\textit{[To be written. Speedrunning evaluation metrics and qualitative results for baseline / baseline+AutoEvolve / PokeAgent.]}

\subsubsection{Sophisticated Supervisor}\label{sec:supervisor_results}

\textit{[To be written. Speedrunning evaluation metrics and qualitative results for supervisor configuration.]}

\subsubsection{Bootstrapped Transfer Learning}\label{sec:transfer_results}

\textit{[To be written. Speedrunning evaluation metrics and qualitative results for transfer configuration.]}

\subsection{Discussion}\label{sec:autoevolve_discussion}

\textit{[To be written. How did AutoEvolve compare to simplest baseline and fully custom PokeAgent? Main failure/success cases. Implications for extending to other domains. Literature-supported analysis.]}
