% ============================================================================
% Chapter 6: Outlook (Headers Only)
% ============================================================================

\newpage
\section{Outlook}\label{sec:outlook}

\subsection{Conclusion}\label{sec:conclusion}

\textit{[To be written. Summary of what was developed: (1) speedrunning benchmark, (2) custom VLM-agent architecture with powerful abstractions for embodied settings, (3) AutoEvolve algorithm for self-adaptation of model harnesses via experiential learning. What was shown and why it is promising.]}

\subsection{Future Work}\label{sec:future_work}

\tersoo{Forward-reference notes from Chapter 4:
 - LLM+RL hybrid scalability: SPD-style approaches where LLMs provide task structure priors
   and RL refines execution; how does this scale beyond the tutorial to long-horizon settings?
 - SLAM/navigation without hand-crafted pathfinding: learned spatial representations,
   SLAM-like map construction from partial observations, reducing dependence on porymap data.
 - Domain-tailored abstractions for long-horizon RPG play: general-purpose coding harnesses
   (Claude Code, Gemini CLI) show promise but lack the operational granularity needed for
   reliable navigation and multi-phase quest management.}

\subsubsection{Model Distillation}\label{sec:distillation}

\textit{[To be written. Training open-source models on a harness learned by a more powerful model; teacher-model paradigm via AutoEvolve; implications for industry.]}

\subsubsection{Full Game Evaluations}\label{sec:full_game}

\textit{[To be written. Extending evaluations beyond three gyms to full game completion.]}

\subsubsection{Extensions to Other Domains}\label{sec:extensions}

\textit{[To be written. Generalization of AutoEvolve to coding, robotics, and other embodied/agentic domains.]}
